---
title: "R bases document"
author: "Aigoin Emilie - Thomas Anne-Laure"
date: "2024-10-09"
output:
  html_document:
    code_folding: hide
    df_print: paged
  pdf_document: default
---
<br>

# Setting up data

### Define the path of the file

```{r}

# Path of the file where the data is saved
folderPathDAT <- "../DAT"

# Path of the file where the results will be saved
folderPathRES <- "../RES"
```

### Load the different packages

```{r, warning = FALSE, message = FALSE}

# Package for data processing, visualization and analysis
library(tidyverse)         

# Package for creating graphics
library(ggplot2)       

# Package for manipulating data frames
library(dplyr)

library(tidyr)

library(gridExtra)

```

### Open the file containing the data

```{r}

# Specify the name of the file to open
filePath <- file.path(folderPathDAT,"/Datagenus.csv")

# Open file with semicolon separations
allData <- read.csv(filePath, sep = ";", header = TRUE)

# Transform all character columns into numerical values
allData <- allData %>% mutate_if(is.character,~ as.numeric(.))

# Select specific columns :
data <- allData %>% select(1:28,          # Counting variables for 27 tree species
                           forest,        # Forest type
                           geology,       # Geological type
                           surface)       # Surface area

# Delete the 1001 row that doesn't match the other rows (not a species fo tree)
data <- data[-1001, ]     

```

### Data verification

##### Outlier detection

```{r}

# Verify empty celluls
colSums(is.na(data))  

# Show where empty cells are
which(is.na(data), arr.ind = TRUE)

# Make a boxplot graph to visually detect outliers
boxplot(data[,2:28], main = "Detection of outliers values", outline = TRUE)

```

##### Locate and remove outliers

```{r}

# If you want to remove the outliers, uncomment the part below

# # Identify the index of the line containing the maximum value of gen10
# max_index <- which.max(data$gen10)
# 
# # Display the line containing the outlier
# data[max_index, ]
# 
# # Delete the line containing the outlier
# data <- data[-max_index, ]

```

##### Set up datasets for our analyses

```{r}

# Data containing only generations
data_gen <- data[,2:28]

# Data containing only generations and forest types
data_gen_forest <- data[,2:29]

# Data containing only generations and geology types
data_gen_geology <- data[,2:30]
data_gen_geology <- data_gen_geology[, -c(28)]

# Data containing only generations and surfaces
data_gen_surface <- data[,-c(1, 29, 30)]

```

### Graphs

##### Graph of forest type

```{r}

data_long <- data_gen_forest %>%
  pivot_longer(cols = starts_with("gen"), # toutes les colonnes des espèces
               names_to = "gen",
               values_to = "Comptage")

desired_order <- c("gen1", "gen2", "gen3", "gen4", "gen5", "gen6", "gen7",
                   "gen8", "gen9", "gen10", "gen11", "gen12", "gen13",
                   "gen14", "gen15", "gen16", "gen17", "gen18", "gen19",
                   "gen20", "gen21", "gen22", "gen23", "gen24",
                   "gen25", "gen26", "gen27")

ggplot(data_long, aes(x = forest, y = Comptage, fill = gen)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(x = "Types de forêts",
       y = "Nombre d'arbres") +
  scale_fill_manual(values = rainbow(27),
                    breaks = desired_order,
                    labels = desired_order) +
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5))  # Centrer le titre

```

##### Graph of geology type

```{r}

data_long <- data_gen_geology %>%
  pivot_longer(cols = starts_with("gen"), # toutes les colonnes des espèces
               names_to = "gen",
               values_to = "Comptage")

ggplot(data_long, aes(x = geology, y = Comptage, fill = gen)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(x = "Types géologiques",
       y = "Nombre d'arbres") +
  scale_fill_manual(values = rainbow(27),
                    breaks = desired_order,
                    labels = desired_order) +
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5))  # Centrer le titre

```

# Part I

### Question 1

##### Density calculation: number of individuals per species / plot area

```{r}

# Creation of a matrix to store the results: stocking densities of per unit area for the 1000 plots
densities <- as.data.frame(matrix(NA,
                                  nrow = nrow(data_gen),
                                  ncol = ncol(data_gen)))

# Loop the plot number (rows) and the tree species (columns), then divide by the corresponding plot area
for (i in 1:nrow(data_gen)) {
  for (j in 1:ncol(data_gen)) {
    densities[i,j] <- data_gen[i,j] / data_gen_surface$surface[i]
  }
}

# Display results
print("Densities are: ")
print(densities)

```

##### Centering and reducing densities

```{r}

# Creation of a matrix to store the results
mean_densities <- as.data.frame(matrix(NA,
                                            nrow = 1,
                                            ncol = 27))

# Sum the densities for each species and divide by 1000 to get the mean
for(j in 1:27) {
  mean_densities[,j] <- sum(densities[,j]) / 1000
}

# Creation of a matrix to store the results
sd_density <- as.data.frame(matrix(NA,
                                            nrow = 1,
                                            ncol = 27))

# Total number of rows (observations)
n <- nrow(densities) 

# Calculate the squared difference between each density and the mean
for (j in 1:27) {
  sum_squared_diff <- 0
  for (i in 1:n) {
  sum_squared_diff <- sum_squared_diff + (densities[i, j] - mean_densities[,j])^2
  }
  # Compute the standard deviation using the variance formula
  sd_density[, j] <- sqrt(sum_squared_diff / (n - 1))
}

# Creation of a matrix to store the results
center_reduct_densities <- as.data.frame(matrix(NA,
                                            nrow = 1,
                                            ncol = 27))

# Subtract the mean and divide by the standard deviation for each species
for (i in 1:n) {
  for (j in 1:27) {
    center_reduct_densities[i,j] <- (densities[i,j] - mean_densities[,j])/(sd_density[,j])
  }
}

# Display results
print("The center reduct densities are: ")
print(center_reduct_densities)

```

##### Calculating barycenters

```{r}

# Initialize the variable 'barycenter' to store the result of the barycenter calculation
barycenter <- 0

# Calculate the barycenter by summing the centered and reduced densities for each species
# The sum is divided by 27 to obtain the average (barycenter)
for (j in 1:27) {
  barycenter <- (1/27) * sum(center_reduct_densities[,j])
}

# Display results
print("Barycenters is: ")
print(barycenter)

```

##### Inertia calculation

```{r}

# Creation of a matrix to store the results
var <- as.data.frame(matrix(NA, 
                             nrow = 1, 
                             ncol = ncol(center_reduct_densities)))

# Calculate the variance for each column
for (j in 1:ncol(center_reduct_densities)) {
  # Initialize the sum of squared differences
  sum_squared_diff <- 0      
  # Calculate the average for column j
  mean_value <- mean(center_reduct_densities[, j])
  
  # Browse each line to calculate the sum of squared differences
  for (i in 1:nrow(center_reduct_densities)) {
    sum_squared_diff <- sum_squared_diff + (center_reduct_densities[i, j] - mean_value)^2
  }
  
  # Calculate the variance for column j
  var[, j] <- (1 / (nrow(center_reduct_densities) - 1)) * sum_squared_diff
}

# Calculate total inertia
inertia_tot <- sum(var)

# Display results
print("Variance is: ")
print(var)

print("Inertia is: ")
print(inertia_tot)

```

### Question 2

##### Calculating the weights of seven forest types

```{r}

# Add forest columns to our dataset
densities$forest <- data$forest
center_reduct_densities$forest <- data$forest

# Calculate the weights
weights=prop.table(table(data_gen_forest$forest))

# Display results
print("Weights are: ")
print(weights)

```

##### Calculation of the barycenters of the seven forest types

```{r}

# Calculation of barycenters for each forest type (using matrices)
barycenter=sapply(unique(data_gen_forest$forest), function(type){
  colMeans(as.matrix(center_reduct_densities[center_reduct_densities$forest == type, 1:27]))
})

# Display results
print("Barycenters are: ")
print(barycenter)

```

##### Calculating square Euclidean norms for each barycenter

```{r}

# Creation of a matrix to store the results
norms_squared <- data.frame(matrix(NA, 
                                 ncol = 7, 
                                 nrow = 1))

# Calculate the squared Euclidean norm for each barycenter
for (j in 1:7) {
  # Calculer la norme euclidienne au carré
  norms_squared[,j] <- sum(barycenter[,j]^2)
}

# Display results
print("Norms squared are: ")
print(norms_squared)

```

##### Calculating inertia

```{r}

# Calculate inter-type inertia
inertia_inter <- sum(weights * norms_squared)

# Display results
print("Inertia inter-types is: ")
print(inertia_inter)

```

##### Calculate the R^2

```{r}

# Calculate the determination coefficient
R2 <- inertia_inter / inertia_tot

# Calculate the information percentage
informations_gain <- R2 * 100

# Display results
print("Gain of information percentage is: ")
print(informations_gain)

```

### Question 3

##### R2 of type-related species

```{r}

# Averaging function
mean_q3 <- function(x) {
  sum(x) / length(x)
}

# Variance calculation function
variance_q3 <- function(x) {
  n <- length(x)
  mean_x <- mean_q3(x)
  sum((x - mean_x)^2) / n
}

# R² calculation for each species
R2_species <- apply(center_reduct_densities[, 1:27], 2, function(x) {
  total_variance <- variance_q3(x)
  
  # Calculation of inter-type variance for each variable
  inter_variance <- sum(sapply(unique(center_reduct_densities$forest), function(type) {
    mean_x_type <- mean(x[center_reduct_densities$forest == type])
    weight <- length(x[center_reduct_densities$forest == type]) / length(x)
    weight * (mean_x_type - mean_q3(x))^2
  }))
  
  inter_variance / total_variance
})

# Order R2_species vector in descending order
R2_species_ordered <- sort(R2_species, decreasing = TRUE)

# Display results
print("R2 by species are: ")
print(R2_species_ordered)

```

##### R2 type-related densities

```{r}

# Calculate the total global variance for each species (columns 1 to 27)
total_variance_global <- apply(center_reduct_densities[, 1:27], 2, variance_q3)

# Calculate the global inter-group variance for the different forest types
inter_variance_global <- sum(sapply(unique(center_reduct_densities$forest), function(type) {
  # Compute the mean for each species for a given forest type
  mean_by_type <- colMeans(center_reduct_densities[center_reduct_densities$forest == type, 1:27])
  # Calculate the weight of each forest type (proportion of the type in the dataset)
  weight <- nrow(center_reduct_densities[center_reduct_densities$forest == type, ]) / nrow(center_reduct_densities)
  # Calculate the weighted variance for this forest type
  weight * sum((mean_by_type - colMeans(center_reduct_densities[, 1:27]))^2)
}))

# Compute the R² of the partition (ratio of inter-group variance to total variance)
R2_partition <- inter_variance_global / sum(total_variance_global)

# Check the correspondence between the partition R² and the arithmetic mean of the species R² values

# Calculate the sum of R² values for the species
sum_R2_species <- sum(R2_species)

# Multiply the partition R² by the number of species to get the mean
mean_R2_species <- sum_R2_species / length(R2_species)

# Display results
print("R2 partition is: ")
print(R2_partition)

# Check if the two R² values are close enough
if (abs(mean_R2_species - R2_partition) < 1e-10) {
  cat("Les deux valeurs de R² sont égales.")
} else {
  cat("Les deux valeurs de R² sont différentes.")
}

```

# Part II

### Creation of matrix

```{r}

# Matrix X: centered-reduced densities (1000 X 1000)
X <- center_reduct_densities[,1:27]
X <- as.matrix(X)

# Matrix Y: forest type indicators (1000 x 7)
n_Y <- length(data$forest)            # Number of observations
k_Y <- length(unique(data$forest))    # Number of distinct categories

# Initialization of the indicator matrix

Y <- matrix(0, 
            nrow = n_Y, 
            ncol = k_Y)

# Construction of the indicator matrix
for (i in 1:n_Y) {
  for (j in 1:k_Y) {
    if (data$forest[i] == j) {
      Y[i, j] <- 1
    }
  }
}

# Z matrix: soil indicators (1000 x 6)

n_Z <- length(data$geology)             # Number of observations
k_Z <- length(unique(data$geology)) + 1 # Number of distinct categories
# Caution! Because no variable 4 but we still have to count it to avoid shifting our data

# Initialization of the indicator matrix
Z <- matrix(0, 
            nrow = n_Z, 
            ncol = k_Z)

# Construction of the indicator matrix
for (i in 1:n_Z) {
  for (j in 1:k_Z) {
    if (data$geology[i] == j) {
      Z[i, j] <- 1
    }
  }
}

Z <- Z[,-4]

# Weight matrices
n <- nrow(X)
p <- ncol(X)

# Matrix W: individual weights (1000 x 1000)
W <- diag(1 / n, n, n)

# Matrix M: variable weights (27 x 27)
M <- diag(1 / p, p, p)

```

### Creation of functions to calculate projections and traces

```{r}

# Functions for calculating projections
project_matrix <- function(A, B) {
  B %*% solve(t(B) %*% B) %*% t(B) %*% A
}

# Function for calculating traces
tr <- function(P) { 
  sum(diag(P)) 
}

```

### Question 1

##### Calcul of projection of Y

```{r}

# Calculation of forest type projections
proj_Y <- Y %*% solve(t(Y) %*% Y) %*% t(Y)

# # Display results
# print("The projection matrix of Y is: ")
# print(P_Y)

```

##### Calculation of the projection of X on the j-th column

```{r}

# Calculating the projections of each x^j onto the space of X
proj_xj <- list()

# Calculate the projection of each column x^j onto the entire space X
for (j in 1:p) {
  proj_xj[[j]] <- project_matrix(X, X[, j, drop = FALSE])
}

# # Display results
# print("The projection of X on the j-th column is: ")
# print(proj_X_xj)

```

##### Calculation of the projection of X on the j-th column and the projection of Y

```{r}

# Initialize a vector to store traces
traces_proj_xj_proj_Y <- numeric(p)

# Creation of a matrix to store the results
for (j in 1:p) {
  proj_xj <- project_matrix(X, X[, j, drop = FALSE])
  
  # Projection of x^j on Y
  proj_xj_proj_Y <- project_matrix(proj_xj, Y)
  
  # Calculation of the trace of projection of x^j and projection of Y
  traces_proj_xj_proj_Y[j] <- tr(proj_xj_proj_Y)
}

# Display results
print("The projection of X on the j-th column and the projection of Y is: ")
print(traces_proj_xj_proj_Y)

```

##### Calculation of the R trace of the projection onto Y

```{r}

# Calculation of R
R <- X %*% M %*% t(X) %*% W

# Calculation of tr(R * Pi_Y)
trace_R_proj_Y <- tr(R %*% proj_Y)

# Display results
print("The R trace of the projection onto Y is: ")
print(trace_R_proj_Y)

```

### Question 2

##### Calculation of each traces of projection of x^j and projection of Z

```{r}

# Initialize a vector to store traces
traces_proj_xj_proj_Z <- matrix(NA, 
                       nrow = p, 
                       ncol = 1)

# Calculation of the projection of x^j onto Z
for (j in 1:p) {
  proj_xj_proj_Z <- project_matrix(X[, j, drop = FALSE], Z)
  
  # Calculation of the trace of projection of x^j and projection of Z
  traces_proj_xj_proj_Z[j] <- tr(proj_xj_proj_Z)
}

# Display results
print("Traces of each projection of x^j and projection of Z are: ")
print(traces_proj_xj_proj_Z)

```

##### Calculation of trace of R and the projection of Z

```{r}

# Compute the matrix product of R with the projection matrix of Z
R_proj_Z <- R %*% project_matrix(Z, Z)

# Calculate the trace of the resulting matrix R_PZ
trace_R_proj_Z <- tr(R_proj_Z)

# Display results
print("The trace of R and the projection of Z is: ")
print(trace_R_proj_Z)

```
